---
title: "ATM KO HS/Npt Phenotype QA/QC"
author: "Daniel M. Gatti, Ph.D."
date: "8/22/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(qtl2)
library(readxl)
library(survival)
library(survminer)
library(tidyverse)

base_dir   = '/media/dmgatti/data1/ColoState/ATM'
pheno_dir  = file.path(base_dir, 'data', 'phenotypes')
pheno_file = file.path(pheno_dir, 'AT phenotype.xlsx')

hap_dir    = file.path(base_dir, 'haplo_reconstr')
qtl2_dir   = file.path(hap_dir,  'qtl2')
covar_file = file.path(qtl2_dir, 'covar.csv')
```

Read in the data.

```{r read_data}
pheno = readxl::read_xlsx(pheno_file, sheet = 'Sheet1', range = 'A1:R793') %>% 
          rename_with(str_to_lower) %>% 
          rename(mouse    = `mouse #`,
                 coat     =`coat color`,
                 necropsy = `necropsy date`,
                 euth     = `fd/euth`,
                 tumor    = `thymic tumor`) %>% 
          select(-(albino:brown), -`thymus rna`, -tail) %>% 
          mutate(sex = str_to_upper(sex))

covar = read_csv(covar_file) %>% 
          separate(id, into = c('batch', 'mouse'), sep = '\\.') %>% 
          select(-batch, -sample) %>% 
          mutate(mouse = if_else(str_detect(mouse, '^[0-9]'), str_c('A', mouse), mouse)) %>% 
          filter(str_detect(mouse, '^A[0-9]+$'))
```

How many mice are there in the phenotypes?

```{r num_mice}
distinct(pheno, mouse) %>% 
  nrow()
```

How many mice are there in the genotypes?

```{r num_mice_geno}
distinct(covar, mouse) %>% 
  nrow()
```

Are there any duplicates in the phenotype data?

```{r duplicate_mice}
count(pheno, mouse) %>% 
  filter(n > 1) %>% 
  left_join(pheno) %>% 
  arrange(mouse)
```

Yes, there are 4 mice with duplicated rows. I don't see a way of distinguishing A1540, A1855 or A56 becuase both mice are the same sex and coat color. I may be able to match A3752 with a mouse in the genotype data based on coat color or sex. I estimated the sex using the genotype data and this is in the 'covar' file.

I'm removing A1540, A1855 or A56 because I have no way of assigning them the correct genotype. 

```{r remove_dupl}
dupl = count(pheno, mouse) %>% 
         filter(n > 1 & mouse != 'A3752') %>% 
         pull(mouse)
pheno = pheno %>% 
          filter(!mouse %in% dupl)
```

How many mice are in both phenotypes and genotypes?

```{r mouse_intersect}
inner_join(pheno, covar, by = 'mouse') %>% 
  nrow()
```

Which mice are not in the intersection?

```{r mouse_setdiff}
anti_join(pheno, covar, by = 'mouse')
```

There are 86 mice that were not genotyped. 



Compare the sex in the phenotype file with sex from the genotypes.

```{r check_sex}
sex_check = pheno %>% 
              select(mouse, sex) %>% 
              rename(sex_p = sex) %>% 
              full_join(covar, by = 'mouse') %>% 
              rename(sex_g = sex) %>% 
              mutate(sex_ok = sex_p == sex_g)
```

Report the number of samples that match, don't match and are NA.?

```{r check_sex2}
count(sex_check, sex_ok)
```

The NA samples are samples that were not genotyped.  Which samples have sex mismatches?

```{r check_sex3}
sex_check %>% 
  select(-sex_p) %>% 
  filter(!sex_ok) %>% 
  left_join(pheno) %>% 
  arrange(mouse)
```

I can resolve A3752 since there is a male sample in the genotype data. I don't see a way to resolve the other samples. They don't have consecutive IDs that might suggest swapping in a tube or plate well.

```{r remove_sample_mismatches}
remove = sex_check %>% 
           select(-sex_p) %>% 
           filter(!sex_ok) %>% 
           left_join(pheno) %>% 
           filter(mouse != 'A3752') %>% 
           pull(mouse)

pheno = pheno %>% 
          filter(!mouse %in% remove) %>% 
          filter(!(mouse == 'A3752' & sex == 'F'))
```

Verify that the sex matches now or that the samples aren't in the genotype data.

```{r veryfy_sex}
pheno %>% 
  select(mouse, sex) %>% 
  rename(sex_p = sex) %>%
  left_join(select(covar, mouse, sex_g = sex), by = 'mouse') %>% 
  mutate(sex_ok = sex_p == sex_g) %>% 
  count(sex_ok)
```

There are 86 samples in the genotypes that aren't in the phenotypes. 

Remove mice with no necropsy date.

```{r remove_no_necropsy}
pheno = pheno %>% 
         filter(!is.na(necropsy))
```

We now have `r nrow(pheno)` samples. 

I will subset the phenotypes later for mapping. But first, I want to make survival plots. There are some notes in the data file about sick and fighting mice, but I'm not sure how to evaluate those. The study is looking for survival due to thymic tumors. I'm not sure how to handle mice that died without thymic tumors.

```{r}
pheno_surv = pheno %>% 
               select(mouse, days, sex, euth, tumor) %>% 
               mutate(event = tumor == 'yes')
```

I don't see any survival difference between sexes.

```{r surv_by_sex}
mod = survfit(Surv(days, event) ~ sex, data = pheno_surv)
ggsurvplot(mod, data = pheno_surv)
```

```{r surv_by_sex}
mod = survfit(Surv(days, event) ~ 1, data = pheno_surv)
ggsurvplot(mod, data = pheno_surv)
```

Organize the brain phenotypes.

```{r count_brain}
pheno = pheno %>% 
          mutate(brain = str_to_upper(brain))
count(pheno, brain)
```

I'm going to create 5 ordered categories based on the categories above: N, AB, CL, CL+, CL++, CL+++. 

```{r create_brain_categories}
pheno = pheno %>% 
          mutate(brain = if_else(str_detect(brain, '---'), NA_character_, brain),
                 brain = str_replace(brain, '\\?', ''),
                 brain = if_else(brain == 'CL++;AB', 'CL++', brain))
count(pheno, brain)
```

Write out the cleaned phenotype data.

```{r write_pheno}
pheno = pheno %>% 
          filter(mouse %in% covar$mouse)
write_csv(pheno, file = file.path(pheno_dir, 'atmko_hs_phenotypes_cleaned.csv'))
```



